---
title: "Assignment 1"
output:
  word_document: 
    highlight: haddock
    fig_caption: yes
    keep_md: yes
  pdf_document: default
date: '`r Sys.Date()`'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 4,fig.height = 4,message = FALSE,
                      warning = FALSE,
                      device = "")
```



```{r library, include = FALSE}
# Install all the necessary packages
load.lib<-c("haven", "dplyr","ggplot2","readr","ggpubr","extrafont","codebook","labelled","psych","splines2","tibble","poputils")

install.lib<-load.lib[!load.lib %in% installed.packages()]
for(lib in install.lib) install.packages(lib,dependencies=TRUE)
sapply(load.lib,require,character=TRUE)

```


```{r col, include = FALSE}
# assign a path, where all files are held and will be saved 

path = "C:/Users/yulleo/OneDrive - Karolinska Institutet/Desktop/Therese/"

# assign the same colors to all graphs
cols <- c("#F8766D", "#00BA38", "#619CFF") 

```

#### Question 1 
Read in the data file on number of colon cancer cases (the file *cases.RData*)
and make sure that you understand the variables included.

```{r q1, results = FALSE}
# Read in R data set

load(paste0(path,"cases.RData"))

# Look at the data set:
View(cases)

# Look at descriptive statistics for the data set:
describe(cases)

# Look at the first rows of the data set:
head(cases)

# Create a codebook for the data set (not necessary for this exercise)

# codebook(cases,survey_overview = FALSE,
#          detailed_scales = FALSE,
#          detailed_variables = FALSE,
#          missingness_report = FALSE,
#          metadata_json = FALSE)

```

**Create graphs showing the number of cases by sex and by age groups.**

Describe what you can conclude from the graphs.

```{r}
# Aggregate the data set by sex 
totals <- cases %>% group_by(sex) %>% 
  summarize(value = sum(n))

cases_sex <- ggplot(totals, aes(x = sex, y = value,fill = sex)) + 
  geom_col() + 
  # define a general theme for the graph
  theme_classic() +
  # make some changes in the theme
  theme(
    legend.position = "none",
    plot.margin = unit(c(0.5,0,0.5,0),"cm"), #define margins for the plot
    axis.title.x = element_blank(),          # take away x-axis title 
    axis.text.x = element_text(size = 10),   # define size for x-axis values   
    axis.title.y = element_text(size = 10),  # define size for y-axis title
    axis.text.y = element_blank(),           # take away y-axis values 
    axis.ticks.y = element_blank())  +         # take away y-axis ticks 
  labs(y = "number of cancer cases")+        # define y-axis title
  scale_fill_manual(values=cols) +          # assign manually colors to the graph
  # include y-values inside the bars
  geom_text(data = totals, aes(x = sex, y = value, label = value),position = position_stack(vjust = 0.5))

cases_sex

```

**Create graphs by sex and by age groups**

```{r}
# Calculate # of cases by sex and age group
totals_age <- cases %>%
  group_by(sex,agegroup) %>%
  summarize(value = sum(n))

cases_age <- ggplot(totals_age, aes(x = agegroup, y = value, fill = sex)) + 
  geom_col() + 
  theme_classic() +
  theme(
    legend.text = element_text(size=7),
    legend.position = "inside",
    legend.position.inside = c(0.1,0.8),
    plot.margin = unit(c(0.5,0,0.5,0),"cm"),
    axis.title.x = element_blank(),
    axis.text.x = element_text(size = 10, angle = 45,vjust=0.3, hjust=1),
    axis.ticks.x = element_blank(),
    axis.title.y = element_text(size = 10),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank())+
  labs(x = "5-year age groups",y = "number of cancer cases")+ 
  scale_fill_manual(values=cols) +
  geom_text(data = totals_age, aes(x = agegroup, group = sex,y = value, label=value),position = position_stack(vjust = 0.5),size = 2)

cases_age

```


#### Question 2
Create a variable for the total number of cases in each calendar year,
separately for males and females. Create graphs showing the number
of cases over calendar years. Describe what you can conclude from the
graphs.


```{r q2, fig.width=8}
# Calculate # of cases by sex and calendar year
totals_year <- cases %>%
  group_by(sex, year) %>%
  summarize(value = sum(n))


cases_year <- ggplot(totals_year, aes(x = as.factor(year), y = value, fill = sex)) + 
  geom_col() + 
  theme_classic() +
  theme(legend.text = element_text(size=7))+
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.1,0.8),
    plot.margin = unit(c(0.5,0,0.5,0),"cm"),
    axis.title.x = element_blank(),
    axis.text.x = element_text(size = 7,family = "Arial", angle = 45,vjust=0.3, hjust=1),
    axis.title.y = element_text(size = 7,family = "Arial"),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.ticks.x = element_blank())+
  labs(x = "5-year age groups",y = "number of cancer cases")+ 
  scale_fill_manual(values=cols) +
  geom_text(aes(group = sex, label=value),position = position_stack(vjust = 0.5),size = 2,angle = 90)

cases_year


```


#### Question 3:
Merge on information on number of persons at risk (the file *population.RData*) in each year, for each age group and sex. Make sure that you
understand the variables included. Does the population file include the
same age groups and calendar years as the file including the number of
cases? Why can this be important to check? 

```{r, results=FALSE}
# Read in population data set:

load(paste0(path,"population.RData"))
describe(population)
head(population)

```


```{r}

# Merge two data sets on sex, age group and year:
joined_df <- left_join(cases,population)

head(joined_df)

```


#### Question 4:
Create a new variable for the incidence rate of colon cancer by dividing
the number of cases with the population size. Describe shortly what an
incidence rate is, and your thoughts on if this is an appropriate way of
calculating an incidence rate.


```{r}

joined_df <- joined_df %>% mutate(ir = n / n_pop, 
                                  ir_extra = ir*10000)

```

#### Question 5:
Collapse the joined data set by sex and calendar year. Create a ne variable of incidence rate. Plot the incidence rate of colon cancer over calendar time, separately by males and females. Describe what you can
conclude from the graphs.


```{r}
# Create a variable of total # of cancer cases by calendar year and sex
joined_df_collapsed <- joined_df %>% group_by(sex,year) %>% 
  summarise(total_pop = sum(n_pop), total_cases = sum(n))

# Create IR per 1 person & per 10,000 people:
joined_df_collapsed <- joined_df_collapsed %>% 
  mutate(ir = total_cases /total_pop) %>% 
  mutate(ir_extra = ir*10000)

ir_year <- ggplot(joined_df_collapsed, aes(x = year, y = ir_extra, color = sex)) + 
  geom_point() + 
  theme_classic() +
  theme(
    legend.text = element_text(size=7),
    legend.position = "inside",
    legend.position.inside = c(0.15,0.8),
    plot.margin = unit(c(0.5,0,0.5,0),"cm"),
    axis.title.x = element_text(size = 10),
    axis.text.x = element_text(size = 10, angle = 45, hjust=1),
    axis.ticks.x = element_blank(),
    axis.title.y = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    axis.ticks.y = element_blank())+
  scale_y_continuous(limits = c(0,NA),n.breaks = 10)+
  labs(x = "calendar year",y = "incidence rate per 10,000") + 
  scale_color_manual(values=cols) 

ir_year

# There is an increasing incidence rate for men and women across years

```


#### Extra:

Create a graph of incidence rates by calendar year, sex for age groups [35-39], [70-74] & [84-89]

```{r}

temp <- joined_df %>% filter(agegroup == "35-39" | agegroup == "70-74" | agegroup == "85-89")

ir_year_age <- ggplot(temp, aes(x = year, y = ir_extra, color = sex)) + 
  geom_point() + 
  theme_classic() +
  theme(
    legend.text = element_text(size=7),
    legend.position = "bottom",
    plot.margin = unit(c(0.5,0,0.5,0),"cm"),
    axis.title.x = element_text(size = 10),
    axis.text.x = element_text(size = 10, angle = 45, hjust=1),
    axis.ticks.x = element_blank(),
    axis.title.y = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    axis.ticks.y = element_blank())+
  scale_y_continuous(limits = c(0,NA),n.breaks = 5)+
  labs(x = "calendar year",y = "incidence rate per 10,000") + 
  scale_color_manual(values=cols) 


plot <- ir_year_age + facet_grid(rows = vars(agegroup), scales = "free") +
  theme(
    strip.background = element_blank(),
    strip.text.x = element_text(
      size = 7, face = "bold.italic")
  )

plot

```


#### Question 6:
Since there is a lot of random variation of the incidence rate from year to
year, we can use a regression model to get smooth estimates of the pattern
of the incidence rate across calendar year. Fit a Poisson model with the
total number of cases as dependent variable, using the population size as
an offset, and calendar year and sex as independent variables.

```{r}
# Create a binary variable female with values 0/1, where female = 0 if sex == F 

joined_df_collapsed <- joined_df_collapsed %>% mutate(female = ifelse(sex == "Female", 0,1))


m1 <- glm(total_cases ~ female + year, family="poisson"(link = "log"), offset = log(total_pop), data=joined_df_collapsed)
# Look at the output:
m1
#summary(m1)

```

#### Question 7:
Based on the model output from above, what is the incidence rate in 1970
among males and females? Based on the model output from above, what is
the incidence rate in 2020 among males and females? 


What assumptions have you made regarding how the incidence rate changes over calendar years and what the difference is between males and females?

```{r}

# Create a newdata set for predictions by subsetting joined_df_total:

newdata1 <- joined_df_collapsed %>% filter(year == 1970 |year == 2020) %>% 
  mutate(total_pop = 1)  

# Obtain predicted IR per person 
newdata1$pred_ir <- predict(m1, newdata1, type = "response")

head(newdata1[,c(1,2,5,8)])

# Obtain predicted IR per 10 000:
newdata1 <- newdata1 %>% mutate(total_pop = 10000)
newdata1$pred_ir_1 <- predict(m1, newdata1, type = "response")

# type = response is an equivalent to exp(predict)

head(newdata1[,c(1,2,6,9)])

```


#### Question 8:
Since colon cancer is more common in older age groups, and the age distribution
has changed in the population, we want to estimate age-specific
rates over calendar years, by sex. 


Again fit a Poisson model, but this time with the age-specific number of cases 
as the dependent variable, and age-specific population size as offset, 
and calendar year, age group and sex as independent variables. 

**Make sure to not assume that the pattern across age groups and across calendar year is the same  for males and females.**

```{r}
joined_df <- joined_df %>% mutate(female = ifelse(sex == "Female", 0,1))

m2 <- glm(n ~ female + year + agegroup + female*agegroup + female*year, family="poisson"(link = "log"), offset = log(n_pop), data=joined_df)

m2

#summary(m2)

```

Based on the model output from above, what is the incidence rate in 1970
in age group 70-74 among males and females? Based on the model output
from above, what is the incidence rate in 2020 in age group 70-74 among
males and females?
```{r}

newdata2 <- joined_df %>% filter(year == 1970 |year == 2020) %>% 
  filter(agegroup == "70-74" | agegroup == "35-39") %>%    
  mutate(n_pop = 10000)  %>% arrange(sex, agegroup,year)  

newdata2 <- joined_df %>% filter(year == 1970 |year == 2020) %>% 
  mutate(n_pop = 10000) %>% arrange(sex, agegroup,year)  


# Obtain predicted IR per person 
newdata2$pred_ir <- predict(m2, newdata2, type = "response")

head(newdata2[,c(1,2,3,7,9)])


```

#### Question 9:
Refit the model above using splines for the effect of calendar year and age
group (use the mid point of each age group), and also make sure to not
assume that the pattern across calendar year is the same across age and
sex. 


```{r plot3,fig.width=8}

# Create a mid point for each age group:
joined_df$age_mid <- age_mid(joined_df$agegroup)

m3 <- glm(n ~ female + nsk(year,df=2)+nsk(age_mid,df=2) + female*nsk(year,df=2) + nsk(year,df=2)*nsk(age_mid,df=2), family="poisson"(link = "log"), offset = log(n_pop), data=joined_df)

m3
#summary(m3)

```


Create graphs showing the incidence rate across calendar time for
males and females at ages 50, 70 and 90.

```{r ,fig.width=8}
# Create a new data set for predictions
newdata3 <- tibble(year = rep(c(1970:2022),6), 
                   female=c(rep(c(rep(0,53),rep(1,53)),3)),
                   age_mid = c(rep(50,106),rep(70,106),rep(90,106)),
                   n_pop = rep(10000,318))

# Obtain predicted IR per 10,000 person 
newdata3$pred_ir <- predict(m3, newdata3, type = "response")


newdata3$female <- factor(newdata3$female, levels = c(0,1), labels = c("female","male"))
newdata3$age_mid <- factor(newdata3$age_mid, levels=c(50,70,90),labels = c("50 years","70 years","90 years"))

plot <- ggplot(data = newdata3, aes(x = year, y = pred_ir, color = female)) +
  geom_line() + 
  theme_classic() +
  theme(
    legend.text = element_text(size=7),
    legend.position = "bottom",
    plot.margin = unit(c(0.5,0,0.5,0),"cm"),
    axis.title.x = element_text(size = 10),
    axis.text.x = element_text(size = 10, angle = 45, hjust=1),
    axis.title.y = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    axis.ticks.y = element_blank(),
    axis.ticks.x = element_blank())+
  labs(x = "calendar year",y = "incidence rate per 10,000", color = "sex")+ 
  scale_color_manual(values=cols) 

plot <- plot + facet_grid(cols = vars(age_mid)) +
  theme(
    strip.background = element_blank(),
    strip.text.x = element_text(
      size = 7, face = "bold.italic"))


plot

```

**Extra** Compare with the observed values for age group [70-74] & [84-89]


```{r plot2, fig.width=8}

temp <- joined_df %>% select(sex,year,female, ir,agegroup) %>% mutate(ir = ir*10000) %>% filter(agegroup == "70-74" | agegroup == "85-89")


plot <- ggplot(data = temp, aes(x = year, y = ir, color = sex)) +
  geom_line() + 
  theme_classic() +
  theme(
    legend.text = element_text(size=7),
    legend.position = "bottom",
    plot.margin = unit(c(0.5,0,0.5,0),"cm"),
    axis.title.x = element_text(size = ),
    axis.text.x = element_text(size = 7, angle = 45, hjust=1),
    axis.title.y = element_text(size = 7),
    axis.text.y = element_text(size = 7),
    axis.ticks.y = element_blank(),
    axis.ticks.x = element_blank())+
  labs(x = "calendar year",y = "incidence rate per 10,000", color = "sex")+ 
  scale_color_manual(values=cols) 

plot <- plot + facet_grid(cols = vars(agegroup)) +
  theme(
    strip.background = element_blank(),
    strip.text.x = element_text(
      size = 7, face = "bold.italic")
  )

plot

```



#### EXTRA:
If we want to compare incidence rates between calendar years, we would like to have a summary statistics over all age groups. However, we have to take into account differences in the age distribution between calendar years. Age-standardized rates allow us to make such a comparison. We, first, calculate direct age-standardized incidence rates by taking for "standard" age distribution in the last available calendar year, 2022.  
```{r}
library(epitools)
library(tidyverse)

# Save a standard population, i.e. the number of population in each age group in year 2022:

standard <- population %>%  filter(year == 2022)

# Join our data set with standard values 

temp <- left_join(joined_df, standard, by = c("agegroup", "sex"), copy = TRUE)

# Calculate direct age-stndardized incidence rates for each calendar year and separately for males and females 
dsr <- temp %>% 
   group_by(sex,year.x) %>% 
   summarise(age_adjust = list(ageadjust.direct(count = n,
            pop = n_pop.x, stdpop = n_pop.y))) %>% 
   mutate(age_adjust = map(age_adjust, as.data.frame.list))  %>% 
   unnest

```


Create a graph of age-standardized incidence rates and compare with non-age-standardized graph:


```{r}
# Change the data set in the long format:

df <- pivot_longer(dsr,cols = ends_with("rate"), names_to = "group",values_to = "ir") %>% 
  mutate(ir_extra = ir*10000)

dsr <- ggplot(df, aes(x = year.x, y = ir_extra, color = group)) + 
  geom_line() + 
  theme_classic() +
  theme(
    legend.text = element_text(size = 10),
    legend.position = "bottom",
    plot.margin = unit(c(0.5,0,0.5,0),"cm"),
    axis.title.x = element_text(size = 10),
    axis.text.x = element_text(size = 10, angle = 45, hjust=1),
    axis.ticks.x = element_blank(),
    axis.title.y = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    axis.ticks.y = element_blank())+
  scale_y_continuous(limits = c(0,NA),n.breaks = 10)+
  labs(x = "calendar year",y = "age_standardized ir per 10,000", color = "") + 
  scale_color_manual(values=cols) 

dsr <- dsr + facet_grid(cols = vars(sex)) +
  theme(
    strip.background = element_blank(),
    strip.text.x = element_text(
      size = 7, face = "bold.italic")
  )

dsr


```


Obtain regression standardized incidence rates, i.e. averaged over age: 


```{r}

# Obtain predictions for all age_mid:

stand <- joined_df %>% mutate(n_pop=10000)

stand$pred_ir_1 <- predict(m3,stand,type = "response") 

# Calculate an average pred_ir_1 by sex and calendar year:
stand <- stand %>% group_by(sex,year) %>% 
  summarise(pred_ir_1_mean = mean(pred_ir_1))


```


Create graphs of standardized over age for each calendar year and sex:

```{r}

age_stand <- ggplot(stand, aes(x = year, y = pred_ir_1_mean, color = sex)) + 
  geom_line() + 
  theme_classic() +
  theme(
    legend.text = element_text(size = 10),
    legend.position = "bottom",
    plot.margin = unit(c(0.5,0,0.5,0),"cm"),
    axis.title.x = element_text(size = 10),
    axis.text.x = element_text(size = 10, angle = 45, hjust=1),
    axis.ticks.x = element_blank(),
    axis.title.y = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    axis.ticks.y = element_blank())+
  scale_y_continuous(limits = c(0,NA),n.breaks = 10)+
  labs(x = "calendar year",y = "age_standardized ir per 10,000", color = "") + 
  scale_color_manual(values=cols) 

age_stand

```






